### 2021-02-25 ch/1-introduction.tex
Some improvements

\chapter{Introduction}\label{ch:introduction}
\section{Problem statement}
In order for \ac{ML} systems to be used more widely,
they have to become more trustworthy.
The susceptibility of \acp{ANN} to adversarial attacks has been well documented, but there are other problems as well.
This includes their opaqueness and their general tendency to take shortcuts;
leading to situations where \acp{ANN} do not do `what we meant', but just what they were explicitly told to do.

@@

% \section{Claims and contributions}%
\section{List of publications and author contributions}%
\label{sec:claims-contributions}
This thesis is based on 3 publications, corresponding to \rangechapref{ch:paper1}{ch:paper3}.
The first one is concerned with \emph{label bias} and the other two with \emph{sampling bias}.
\begin{enumerate}
  \item Thomas Kehrenberg, Zexun Chen, and Novi Quadrianto (2020). ``Tuning Fairness by Balancing Target Labels.''
    In: \textit{Frontiers in Artificial Intelligence} 3, p.\ 33.
    \citep{kehrenberg2020tuning}
  \item Thomas Kehrenberg, Myles Bartlett, Oliver Thomas, and Novi Quad\-ri\-an\-to (2020).
    ``Null-sampling for Interpretable and Fair Representations.''
    In: \textit{Computer Vision - ECCV 2020}. Glasgow, UK: Springer International Publishing.
    \citep{kehrenberg2020nullsampling}
  \item Thomas Kehrenberg, Viktoriia Sharmanska, Myles Bartlett, and Novi Quadrianto (2021).
    ``Learning with Perfect Bags: Addressing Hidden Stratification with Zero Labeled Data.''
    \textit{Under review at ICML 2021}.
    \citep{kehrenberg2020zeroshot}
\end{enumerate}
A shorter version of \citet{kehrenberg2020tuning} was published as a workshop paper:
\begin{itemize}
  \item Thomas Kehrenberg, Zexun Chen, and Novi Quadrianto (2018).
    ``Interpretable Fairness via Target Labels in Gaussian Process Models.''
    In: \emph{Workshop on Ethical, Social and Governance Issues in AI at NeurIPS}. Montreal, Canada.
\end{itemize}

@@

>>>>

\chapter{Introduction}\label{ch:introduction}
\section{Problem statement}
In order for \ac{ML} systems to be used more widely,
they have to become more trustworthy.
The susceptibility of \acp{NN} to adversarial attacks has been well documented, but there are other problems as well.
This includes their opaqueness and their general tendency to take shortcuts;
leading to situations where neural networks do not do `what we meant', but just what they were explicitly told to do.

@@

% \section{Claims and contributions}%
\section{List of publications and author contributions}%
\label{sec:claims-contributions}
This thesis is based on 3 publications, corresponding to \rangechapref{ch:paper1}{ch:paper3}.
The first one is concerned with \emph{label bias} and the other two with \emph{sampling bias}.

@@

\subsection{Publication 1}
\begin{refsection}[allreferences]
    \nocite{kehrenberg2020tuning}
    \printbibliography[heading=none]
\end{refsection}
\noindent A shorter version was published as a workshop paper:
\begin{refsection}[allreferences]
    \nocite{kehrenberg2018interpretable}
    \printbibliography[heading=none]
\end{refsection}
\noindent\textsc{Contributions:}
\begin{itemize}
  \item I conceived the idea of using Target Labels to target a balanced set. I developed the proof from a starting point that my supervisor pointed me to. I wrote all of the code dealing with the modified loss function and nearly all of the remaining code as well. I ran most of the experiments and wrote all of the methods section and most of the remaining text as well.
  \item Z. Chen was a discussion partner and helped run the experiments, and wrote some parts of the code.
  \item N. Quadrianto suggested the initial direction of the work, was a discussion partner, and helped write the introduction and related work.
\end{itemize}

\subsection{Publication 2}
\begin{refsection}[allreferences]
    \nocite{kehrenberg2020nullsampling}
    \printbibliography[heading=none]
\end{refsection}%
\noindent\textsc{Contributions:}
\begin{itemize}
  \item I conceived the idea of using an \acf{INN} and a representative set to learn an invariant representation. I wrote a large part of the code, and ran about half the experiments. I wrote a significant part of the text.
  \item M. Bartlett developed a large part of the tricks needed for training the \ac{INN} successfully. He wrote a significant part of the code, and of the text.
  \item O. Thomas helped with writing the code and with running the experiments.
  \item N. Quadrianto gave feedback on the progress and suggested directions to explore.
\end{itemize}

\subsection{Publication 3}
\begin{refsection}[allreferences]
    \nocite{kehrenberg2020zeroshot}
    \printbibliography[heading=none]
\end{refsection}%
\noindent\textsc{Contributions:}
% -  perfect bags
%     - I
%         - developed the idea of distribution matching
%         - developed the idea of clustering an auxiliary set
%     - Vika
%         - head the idea for how to frame the method
%         - worked on introduction/related work/method
%     - Myles
%         - ran experiments
%         - came up with NN architecture
%         - came up with idea of bags-of-batches
%     - Novi
%         - combined the two pieces that I had into a whole
\begin{itemize}
  \item I developed the idea of distribution matching as an extension of the previous paper. In addition, I tried to achieve similar goals by applying clustering to
\end{itemize}

### 2021-02-25 ch/2-related-work.tex
Some improvements

\chapter{Related work}\label{ch:related-work}
%
% =========================
%      old introduction
% =========================
% \section{Introduction}\label{introduction}
% Fairness in Machine Learning is a topic that has recently received increasing attention.
% Machine Learning has become very widespread in our society
% which brings some dangers with it.
% Decisions such as credit eligibility and hiring are increasingly done automatically.
% However, the algorithms that are used in these decisions can have strong biases
% that stem from the training data.
% Often, the algorithms are even more biased than the training data.
% \citet{kamishima2011fairness} give a great illustration for this
% (adapted from \citet{calders2010three}).
% The illustration is based on the Adult/Census Income dataset \citep{kohavi1996scaling}
% which contains records of individuals with thirteen attributes, among them age and gender.
% The task is to predict whether or not the individual earns more than \$50K per year.
% In the dataset, 32\% of the men have a positive label (\ie, they earn more than \$50K),
% but of the women only 11\% have a positive label.
% With this ratio of about 3:1, the dataset is already skewed.
% However, when training a \href{https://en.wikipedia.org/wiki/Naive_Bayes_classifier}{Na\"ive Bayes classifier} on the data,
% the ratio in the predictions is even more unfair:
% the amount of women who get a positive prediction is even lower than 11\%.
% If the Na\"ive Bayes classifier does not get access to the gender information,
% the predictions will be a bit fairer,
% but they are still more unfair than the original dataset labels.
%
% This phenomenon of predictions being more unfair than an already skewed dataset
% is due to the generalization that classifiers have to make.
% It is easier to classify all but the most outstanding women as low income
% than to learn a complicated decision boundary.
% Furthermore, removing the gender information does not solve the problem
% because the gender can be inferred from other data.
%
% The following is a review of the literature on Fairness in Machine Learning.
% Several different approaches to and definition of Fairness are presented.
% The focus is on identifying the major trends in the field from the beginning.
\section{\NoCaseChange{Two views of the dataset bias problem}}\label{two-views-of-the-dataset-bias-problem}
As hinted to in the introduction,
there is a divide in the relevant literature in how the problem of dataset bias is approached.
On the one hand, there is the ``groundtruth-centric'' view -- expressed through most of the introduction --
that the training set is biased but the deployment setting is not.
By \emph{deployment setting} I am here referring to the real distribution
that will be encountered by the machine learning system once it is in use.

@@

This method can be improved further by taking into account the labels
when constructing the fair representation.
If this is not done, \(z\) loses the ranking information from \(x\) (see \citet{feldman2015certifying} above).
To this end, a second latent variable is introduced, \(\tilde{z}\),
which encodes the variation in \(z\) that is not explained by the class labels \(y\).
\(z\) is then determined by \(y\) and \(\tilde{z}\),
and \(x\) is determined by \(z\) and \(s\) as before.
\(\tilde{z}\) and \(s\) have independent priors.
This structure ensures that \(y\) can be predicted from \(z\).
However, this introduces a new problem:
if \(y\) is correlated with \(s\), then \(z\) will be as well.
To overcome this, an additional penalty term is introduced that forces \(P(z|s=0)\)
and \(P(z|s=1)\) to be as close as possible.
This is realized with a measure of distance
between distributions called Maximum Mean Discrepancy (MMD).

The experiments were done on the Adult/Census Income dataset,
the German Credit dataset and the Heritage Health dataset.
To test whether \(s\) can be recovered from \(z\),
a Random Forest model and a Logistic Regression model were trained to predict \(s\) from \(z\).
This is compared to \citet{zemel2013learning}.
The proposed model can outperform \citet{zemel2013learning} on some tasks, but many results seem inconclusive.
Using MMD for an additional unfairness penalty, seems to improve fairness.
When training a classifier on \(z\) to predict \(y\), there is a small drop in accuracy.
In general, it is not clear whether this method gives better performance than Feldman et
al.~However, the algorithm by Louizos seems more general and more widely applicable
as it simply re-uses a widely used technique, variational autoencoders.

\section{Other fairness criteria}%
\label{improved-definitions-of-fairness}
The early literature on fairness in machine learning used only \ac{DP}
and similarity-based fairness to judge the fairness of a classifier.
\citet{kleinberg2016inherent} formalized three different fairness conditions.
The authors consider a scenario where the data points \(x\) are sorted into bins \(b\)
and each bin is associated with a prediction score \(f = P(\hat{y}=1|b)\)
where \(\hat{y}\) is the predicted class label and \(x\) the (non-sensitive) features.
\begin{enumerate}
\item
  Calibration within groups:
  If the prediction score for a given bin is \(f\),
  then when considering all the data points with group \(s\) in the bin,
  a fraction of \(f\) of those should have the class label \(y=1\).
  In other words, the prediction score \(f\) is well calibrated with respect to group \(s\).
  This should be the case for all groups.
\item
  Balance for the negative class:
  The true negative rate should be the same for both groups: \(P(\hat{y}=0|y=0,s=0) = P(\hat{y}=0|y=0,s=1)\).
\item
  Balance for the positive class:
  The true positive rate should be the same for both groups: \(P(\hat{y}=1|y=1,s=0) = P(\hat{y}=1|y=1,s=1)\).
\end{enumerate}
These are formalizations of things that have been proposed as definitions of fairness.
The first criterion essentially ensures that the predictor works correctly for both groups.
This prevents a classifier from predicting one group correctly
but always returning a negative answer for the other group.
The second and third put emphasis on negative and positive class labels respectively.
In criterion 2, we allow the classifier to mis-classify those data points with \(y=0\),
but we want to make sure that the misclassification rate is the same for both groups.
In other words, members of the different groups have the same chance
to get a correct classification if they should receive a negative classification.
The same holds for criterion 3 and positive classifications (\(y=1\)).

>>>>

\chapter{Related work}\label{ch:related-work}
%
% =========================
%      old introduction
% =========================
% \section{Introduction}\label{introduction}
% Fairness in Machine Learning is a topic that has recently received increasing attention.
% Machine Learning has become very widespread in our society
% which brings some dangers with it.
% Decisions such as credit eligibility and hiring are increasingly done automatically.
% However, the algorithms that are used in these decisions can have strong biases
% that stem from the training data.
% Often, the algorithms are even more biased than the training data.
% \citet{kamishima2011fairness} give a great illustration for this
% (adapted from \citet{calders2010three}).
% The illustration is based on the Adult/Census Income dataset \citep{kohavi1996scaling}
% which contains records of individuals with thirteen attributes, among them age and gender.
% The task is to predict whether or not the individual earns more than \$50K per year.
% In the dataset, 32\% of the men have a positive label (\ie, they earn more than \$50K),
% but of the women only 11\% have a positive label.
% With this ratio of about 3:1, the dataset is already skewed.
% However, when training a \href{https://en.wikipedia.org/wiki/Naive_Bayes_classifier}{Na\"ive Bayes classifier} on the data,
% the ratio in the predictions is even more unfair:
% the amount of women who get a positive prediction is even lower than 11\%.
% If the Na\"ive Bayes classifier does not get access to the gender information,
% the predictions will be a bit fairer,
% but they are still more unfair than the original dataset labels.
%
% This phenomenon of predictions being more unfair than an already skewed dataset
% is due to the generalization that classifiers have to make.
% It is easier to classify all but the most outstanding women as low income
% than to learn a complicated decision boundary.
% Furthermore, removing the gender information does not solve the problem
% because the gender can be inferred from other data.
%
% The following is a review of the literature on Fairness in Machine Learning.
% Several different approaches to and definition of Fairness are presented.
% The focus is on identifying the major trends in the field from the beginning.
\section{\NoCaseChange{Two views of the dataset bias problem}}\label{two-views-of-the-dataset-bias-problem}
As hinted to in the introduction,
there is a divide in the relevant literature in how the problem of dataset bias is approached.
On the one hand, there is the ``ground-truth-centric'' view -- expressed through most of the introduction --
that the training set is biased but the deployment setting is not.
By \emph{deployment setting} I am here referring to the real distribution
that will be encountered by the machine learning system once it is in use.

@@

This method can be improved further by taking into account the labels
when constructing the fair representation.
If this is not done, \(z\) loses the ranking information from \(x\) (see \citet{feldman2015certifying} above).
To this end, a second latent variable is introduced, \(\tilde{z}\),
which encodes the variation in \(z\) that is not explained by the class labels \(y\).
\(z\) is then determined by \(y\) and \(\tilde{z}\),
and \(x\) is determined by \(z\) and \(s\) as before.
\(\tilde{z}\) and \(s\) have independent priors.
This structure ensures that \(y\) can be predicted from \(z\).
However, this introduces a new problem:
if \(y\) is correlated with \(s\), then \(z\) will be as well.
To overcome this, an additional penalty term is introduced that forces \(P(z|s=0)\)
and \(P(z|s=1)\) to be as close as possible.
This is realized with a measure of distance between distributions called \acf{MMD}.

The experiments were done on the Adult/Census Income dataset,
the German Credit dataset and the Heritage Health dataset.
To test whether \(s\) can be recovered from \(z\),
a Random Forest model and a Logistic Regression model were trained to predict \(s\) from \(z\).
This is compared to \citet{zemel2013learning}.
The proposed model can outperform \citet{zemel2013learning} on some tasks, but many results seem inconclusive.
Using \ac{MMD} for an additional unfairness penalty, seems to improve fairness.
When training a classifier on \(z\) to predict \(y\), there is a small drop in accuracy.
In general, it is not clear whether this method gives better performance than \citet{feldman2015certifying}.
However, the algorithm by \citet{louizos2016variational} seems more general and more widely applicable
as it simply re-uses a widely used technique, variational autoencoders.

\section{Other fairness criteria}%
\label{improved-definitions-of-fairness}
The early literature on fairness in machine learning used only \ac{DP}
and si\-mi\-la\-ri\-ty-based fairness to judge the fairness of a classifier.
\citet{kleinberg2016inherent} formalized three different fairness conditions.
The authors consider a scenario where the data points \(x\) are sorted into bins \(b\)
and each bin is associated with a prediction score \(f = P(\hat{y}=1|b)\)
where \(\hat{y}\) is the predicted class label and \(x\) the (non-sensitive) features.
\begin{enumerate}
\item
  Calibration within groups:
  If the prediction score for a given bin is \(f\),
  then when considering all the data points with group \(s\) in the bin,
  a fraction of \(f\) of those should have the class label \(y=1\).
  In other words, the prediction score \(f\) is well calibrated with respect to group \(s\).
  This should be the case for all groups.
\item
  Balance for the negative class:
  The true negative rate should be the same for both groups: \(P(\hat{y}=0|y=0,s=0) = P(\hat{y}=0|y=0,s=1)\).
\item
  Balance for the positive class:
  The true positive rate should be the same for both groups: \(P(\hat{y}=1|y=1,s=0) = P(\hat{y}=1|y=1,s=1)\).
\end{enumerate}
These are formalizations of things that have been proposed as definitions of fairness.
The first criterion essentially ensures that the predictor works correctly for both groups.
This prevents a classifier from predicting one group correctly
but always returning a negative answer for the other group.
The second and third put emphasis on negative and positive class labels respectively.
In criterion 2, we allow the classifier to mis-classify those data points with \(y=0\),
but we want to make sure that the misclassification rate is the same for both groups.
In other words, members of the different groups have the same chance
to get a correct classification if they should receive a negative classification.
The same holds for criterion 3 and positive classifications (\(y=1\)).
